Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-12 doc/mc-paper.tex
typos

One approach to model counting, known as {\em knowledge compilation}
is to transform the formula into a structured form for which model
counting becomes more tractable.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
directed acyclic graph, with terminal nodes labeled by Boolean
variables and their complements, and with each nonterminal node
labeled by Boolean operation And or Or.  Restrictions are place on the
structure of the graph such that a count of the models can be computed
by a single bottom-up traversal of the graph.

@@

Fichte, Hecher, and Roland~\cite{fichte:sat:2021} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  We compare the performance of
our tool chain to theirs as part of our experimental evluation in
Section~\ref{sect:experimental}.  Our experience with developing a
formally verified proof system has shown us that, even for the well
understood framework of extended resolution, ensuring that the system
is sound can be challenging.

@@

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
unalterered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CRAT file to create
the appropriate clauses.  Consider an {\em argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
Table~\ref{tab:defining}(A), we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

@@

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rational
numbers we term $\drational$, consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

>>>>

One approach to model counting, known as {\em knowledge compilation}
is to transform the formula into a structured form for which model
counting becomes more tractable.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a Boolean formula as a
directed acyclic graph, with terminal nodes labeled by Boolean
variables and their complements, and with each nonterminal node
labeled by Boolean operation And or Or.  Restrictions are placed on the
structure of the graph such that a count of the models can be computed
by a single bottom-up traversal of the graph.

@@

Fichte, Hecher, and Roland~\cite{fichte:sat:2021} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  We compare the performance of
our tool chain to theirs as part of our experimental evaluation in
Section~\ref{sect:experimental}.  Our experience with developing a
formally verified proof system has shown us that, even for the well
understood framework of extended resolution, ensuring that the system
is sound can be challenging.

@@

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CRAT file.  Some may be
unaltered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CRAT file to create
the appropriate clauses.  Consider an {\em argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
Table~\ref{tab:defining}(A), we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the {\em guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

@@

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rational
numbers we term $\drational$, consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.
